---
title: "Confusion Matrix and Metrics"
author: "Hansjörg Neth, SPDS, uni.kn"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Confusion Matrix and Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Behold the aptly named "confusion matrix":

|              | Condition |          |       |          |            
| ------:      |:----:|:----:|:------:|:----:|              
| **Decision** | present (`TRUE`):      | absent (`FALSE`):     |     Sum:  |  (b) by decision:      |  
| positive (`TRUE`):   | `hi`         | `fa`         | `dec.pos` | `PPV` = `hi`/`dec.pos` |
| negative (`FALSE`):  | `mi`         | `cr`         | `dec.neg` | `NPV` = `cr`/`dec.neg` |
|      Sum:    | `cond.true`  | `cond.false` |       `N` |                        |
| (a) by condition  | `sens` = `hi`/`cond.true` | `spec` = `cr`/`cond.false` | |       |

Ordinary people, but also medical experts and social scientists, inevitably struggle to understand the implications of this matrix. 
This is no surprise when considering explanatory attempts like the corresponding article on [Wikipedia](https://en.wikipedia.org/wiki/Confusion_matrix), which squeezes more than a dozen metrics out of four essential frequencies (`hi`, `mi`, `fa`, and `cr`). While each particular metric is quite simple, their abundance and inter-dependence can be overwhelming.

In the following, we try to disentangle the profusion of measures and summarize the parts of the confusion matrix that a risk-literate person really needs to know.


## Basics

In its most basic form, the confusion matrix looks like this:

|              | Condition | |
| ------:      |:----:|:----:|      
| **Decision** | present (`TRUE`): | absent (`FALSE`): | 
| positive (`TRUE`):     |  `hi`   |  `fa`    |
| negative (`FALSE`):    |  `mi`   |  `cr`    | 


### Adopting 2 perspectives on a population

Essentially, the confusion table views a population of `N` individuals from two different perspectives. Adopting a perspective means that we distinguish between individuals on the basis of some criterion. The two criteria used here are:

(a) each individual's _condition_, which can either be present (`TRUE`) or absent (`FALSE`), and  
(b) each individual's _decision_, which can either be `positive` (`TRUE`) or `negative` (`FALSE`).   

Numerically, the adoption of each of these two perspectives splits the population into two subgroups.^[To split a group into subgroups, some criterion for classifying the individuals of the group has to be used. If a criterion is binary (i.e., assigns only two different values), its application yields two subgroups. In the present case, both an individual's _condition_ and the corresponding _decision_ are binary criteria.] Applying two different splits of a population into two subgroups results in $2 \times 2 = 4$ cases, which form the core of the confusion matrix:

1. `hi` represents _hits_ (or true positives): condition present (`TRUE`) & decision positive (`TRUE`). 
2. `mi` represents _misses_ (or false negatives): condition present (`TRUE`) & decision negative (`FALSE`). 
3. `fa` represents _false alarms_ (or false positives): condition absent (`FALSE`) & decision positive (`TRUE`). 
4. `cr` represents _correct rejections_ (or true negatives): condition absent (`FALSE`) & decision negative (`FALSE`). 

Importantly, all frequencies required to understand and compute various metrics are combinations of these four frequencies --- which is why we refer to them as the four _essential_ frequencies (see the vignette on [Data formats](data_formats.html)). For instance, adding up the columns and rows of the matrix yields the frequencies of the two subgroups that result from adopting our two perspectives on the population `N` (or splitting `N` into subgroups by applying two binary criteria):

(a) by condition (corresponding to the two columns of the confusion matrix): 

$$ 
\begin{aligned}
\texttt{N} \ &= \ \texttt{cond.true} & +\ \ \ \ \ &\texttt{cond.false}   & \textrm{(a)}      \\
           \ &= \ (\texttt{hi} + \texttt{mi}) & +\ \ \ \ \ &(\texttt{fa} + \texttt{cr}) \\
\end{aligned}
$$ 

(b) by decision (corresponding to the two rows of the confusion matrix): 

$$ 
\begin{aligned}
\texttt{N} \ &= \ \texttt{dec.pos} & +\ \ \ \ \ &\texttt{dec.neg}   & \textrm{(b)}      \\
           \ &= \ (\texttt{hi} + \texttt{fa}) & +\ \ \ \ \ &(\texttt{mi} + \texttt{cr}) \\
\end{aligned}
$$ 

### Adopting a 3rd perspective: Accuracy

A third way of grouping the four essential frequencies results from asking the question: Which of the four essential frequencies are _correct_ decisions and which are _erroneous_ decisions? Crucially, this question about decision _accuracy_ can neither be answered by only considering each individual's condition (i.e., the columns of the matrix), nor can it be answered by only considering each individual's decision (i.e., the rows of the matrix). Instead, the question requires considering the correspondence between condition and decision. Checking the correspondence between rows and columns for the four essential frequencies yields an important insight: The confusion matrix contains _two_ types of correct decisions and _two_ types of errors:

- A decision is _correct_, when it corresponds to the condition. This is the case for two cells in the matrix:

    - `hi`: condition present (`TRUE`) & decision positive (`TRUE`).
    - `cr`: condition absent (`FALSE`) & decision negative (`FALSE`).
    
- A decision is _incorrect_ or _erroneous_, when it does not correspond to the condition. This also is the case for two cells in the matrix:

    - `mi`: condition present (`TRUE`) & decision negative (`FALSE`).
    - `fa`: condition absent (`FALSE`) & decision positive (`TRUE`).
    
    
+++ here now +++


(c) by the correspondence of decisions to conditions (corresponding to the two diagonals of the confusion matrix): 

$$ 
\begin{aligned}
\texttt{N} \ &= \ \texttt{dec.cor} & +\ \ \ \ \ &\texttt{dec.err}   & \textrm{(c)}      \\
           \ &= \ (\texttt{hi} + \texttt{cr}) & +\ \ \ \ \ &(\texttt{mi} + \texttt{fa}) \\
\end{aligned}
$$ 


### Beware some sources of confusion

It may be instructive to point out two possible sources of confusion, so that they can be deliberately avoided later:

1. Beware of alternative names of `mi` and `cr`:

    - Misses `mi` are often called "false negatives", but actually are in the `cond.true` column (on the left).
    - Correct rejections `cr`are often called "true negatives", but are in the `cond.false` column (on the right).

Thus, the terms "false" and "true" in the compound cell names refer to the relationship between or correspondence of condition and decision, rather than to the condition alone. To avoid this source of confusion, we prefer the terms `mi` and `cr`. 

2. Beware of `dec.cor` and `dec.err`:  
It may be tempting to refer to instances of `dec.cor` and `dec.err` as "true decisions" and "false decisions". 
However, this would invite conceptual confusion, as "true decisions" actually include `cond.false` cases (`cr`) and "false decisions" actually include `cond.true` cases (`mi`). Again, we prefer the less confusing terms "correct decisions" vs. "erroneous decisions".


## Relationships

Overall, the following relationships hold among the frequencies:

$$ 
\begin{aligned}
\texttt{N} &= \texttt{hi} + \texttt{mi} + \texttt{fa} + \texttt{cr} \\
           &= \texttt{cond.true} + \texttt{cond.false} \\
           &= \texttt{dec.pos} + \texttt{dec.neg}      \\
           &= \texttt{dec.cor} + \texttt{dec.err}      \\
\end{aligned}
$$ 

Different perspectives:

(a) The 2 columns of the confusion table:

$$ 
\begin{aligned}
\texttt{cond.true}  &= \texttt{hi} + \texttt{mi} \\
\texttt{cond.false} &= \texttt{fa} + \texttt{cr} \\
\end{aligned}
$$ 

(b) The 2 rows of the confusion table:

$$ 
\begin{aligned}
\texttt{dec.pos}  &= \texttt{hi} + \texttt{fa} \\
\texttt{dec.neg}  &= \texttt{mi} + \texttt{cr} \\
\end{aligned}
$$ 




## Metrics

Metrics as dependent measures that result from splitting into subgroups. Different ways of describing this result. 

Accuracy as a dependent measure: some combination that describes the relationship between correct and incorrect decisions. 

Different measures emphasize different aspects or have different purposes. 

 
(c) The 2 diagonals of the confusion table:

$$ 
\begin{aligned}
\texttt{dec.cor}      &= \texttt{hi} + \texttt{cr} \\
\texttt{dec.err}      &= \texttt{mi} + \texttt{fa} \\
\end{aligned}
$$ 

Note that it may be tempting to call instances of `dec.corr` and `dec.err` as "true decisions" and "false decisions". 
However, this would invite conceptual confusion, as "true decisions" actually include `cond.false` cases (`cr`) and "false decisions" actually include `cond.true` cases (`mi`).



### Accuracy `acc`

> Accuracy := Proportion or percentage of cases correctly classified.

$$
\begin{aligned}
\texttt{acc} &= \frac{\texttt{hi} + \texttt{cr}}{\texttt{hi} + \texttt{mi} + \texttt{fa} + \texttt{cr}} 
             = \frac{\texttt{dec.cor}}{\texttt{dec.cor} + \texttt{dec.err}} = \frac{\texttt{dec.cor}}{\texttt{N}}  
\end{aligned}
$$


### Weighted accuracy `wacc`

Weighted accuracy: Taking into account the importance of errors.



### Matthew correlation coefficient `mcc`

3. Matthew correlation coefficient (mcc)


## References

Links to Wikipedia articles:

- [Confusion matrix](https://en.wikipedia.org/wiki/Confusion_matrix) (URL: https://en.wikipedia.org/wiki/Confusion_matrix)



## All `riskyr` Vignettes

- [User guide](user_guide.html)
- [Data formats](data_formats.html)
- [Confusion matrix](confusion_matrix.html)

<!-- eof. -->




# Template

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
