---
title: "Risk Literacy Toolbox"
author: "Hansjörg Neth, SPDS, uni.kn"
date: "2017 12 25"
output:
  html_document:
    keep_tex: false
    self_contained: true # true/false # TRUE can cause errors with rmdformats
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    code_folding: show # show/hide
    highlight: kate # textmate default kate haddock monochrome #
    lightbox: true # TRUE by default
    gallery: false
    use_bookdown: false
    fig_width: 10 # in inches
documentclass: article
classoption: a4
fontsize: 11pt
geometry: margin=1.0in
linkcolor: blue
urlcolor: blue
editor_options: 
chunk_output_type: console
---

```{r preamble, eval = TRUE, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
## (1) Housekeeping: 
rm(list=ls()) # clean all.
# cur.path <- dirname(rstudioapi::getActiveDocumentContext()$path)
# setwd(cur.path) # set to current directory
setwd("~/Desktop/stuff/Dropbox/GitHub/riskyr/_code/") # set to current directory
# list.files() # all files + folders in current directory
fileName <- "RiskLiteracyTools_now.Rmd"

## (2) Packages:
library(rmarkdown)
library(knitr)
library(rmdformats)
# library(yarrr)

## (3) Global options:
options(max.print="75")
opts_chunk$set(echo = TRUE,
	             cache = TRUE,
               prompt = FALSE,
               tidy = FALSE,
               collapse = TRUE, # set to TRUE in answers 
               comment = "#>",
               message = FALSE,
               warning = FALSE,
               fig.width = 10, fig.height = 6)
opts_knit$set(width=75)
```

Graphical settings (for `ggplot2`): 

```{r graphic_settings, eval = TRUE, echo = FALSE}
## (1) Packages: 
library(ggplot2)
library(cowplot)
library(RColorBrewer)

## (2) Graphic parameters:
opar <- par() # saves original (default) par settings
par(opar)  # restores original (default) par settings

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ # 
## (3) Colors:
# (a) Using yarrr:
# col.pal <- yarrr::piratepal(palette = "basel",  trans = .5) # [does not work]

## (b) Using RColorBrewer: 
# display.brewer.all()
# cols <- brewer.pal(4, "Set1")
# brewer.pal.info["Reds",]
# brewer.pal.info["Greens",]
# brewer.pal.info["Blues",]
col.reds <- brewer.pal(9, "Reds")
col.greens <- brewer.pal(9, "Greens")

## (c) Using color names:
col.ppv <- "orange3" # firebrick" "red3"
col.npv <- "steelblue4" # "green4" "gray50" "brown4" "chartreuse4"  

## (d) unikn.col palette (as RGB, without transparency): 

## Define base color "seeblau":
seeblau <- rgb(0, 169, 224, names = "seeblau", maxColorValue = 255) # seeblau.4 (non-transparent)

# in one df (for the yarrr package): 
unikn.pal = data.frame(
  "seeblau1" = rgb(204, 238, 249, maxColorValue = 255), #  1. seeblau1 (non-transparent)
  "seeblau2" = rgb(166, 225, 244, maxColorValue = 255), #  2. seeblau2 (non-transparent)
  "seeblau3" = rgb( 89, 199, 235, maxColorValue = 255), #  3. seeblau3 (non-transparent)
  "seeblau4" = rgb(  0, 169, 224, maxColorValue = 255), #  4. seeblau4 (= seeblau base color)
  "black"    = rgb(  0,   0,   0, maxColorValue = 255), #  5. black
  "seegrau4" = rgb(102, 102, 102, maxColorValue = 255), #  6. grey40 (non-transparent)
  "seegrau3" = rgb(153, 153, 153, maxColorValue = 255), #  7. grey60 (non-transparent)
  "seegrau2" = rgb(204, 204, 204, maxColorValue = 255), #  8. grey80 (non-transparent)
  "seegrau1" = rgb(229, 229, 229, maxColorValue = 255), #  9. grey90 (non-transparent)
  "white"    = rgb(255, 255, 255, maxColorValue = 255), # 10. white
  stringsAsFactors = FALSE)

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ # 
## (4) ggplot Themes:  
my.theme <-  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = "12", color = "black", hjust = 0.0),
        axis.title = element_text(face = "plain", size = 11, color = "black"),
        axis.text = element_text(face = "plain", size = 10, color = "gray30"),
        # axis.line = element_line(size = 0.75, color = "black", linetype = 1), 
        axis.ticks = element_line(size = 0.75, color = "gray10", linetype = 1), 
        #panel.background = element_rect(fill = "gray95", color = "gray20"),
        panel.grid.major.x = element_line(color = "gray66", linetype = 1, size = .2),
        panel.grid.major.y = element_line(color = "gray33", linetype = 1, size = .2),
        #panel.grid.minor.x = element_blank(), 
        #panel.grid.minor.y = element_blank(),
        legend.position = "none"
  )

my.theme.legend <- theme_bw() +
  theme(plot.title = element_text(face = "bold", size = "12", color = "black", hjust = 0.0),
        axis.title = element_text(face = "plain", size = 11, color = "black"),
        axis.text = element_text(face = "plain", size = 10, color = "gray30"),
        # axis.line = element_line(size = 0.75, color = "black", linetype = 1), 
        axis.ticks = element_line(size = 0.75, color = "gray10", linetype = 1),   
        #panel.background = element_rect(fill = "gray95", color = "gray20"),
        panel.grid.major.x = element_line(color = "gray66", linetype = 1, size = .2),
        panel.grid.major.y = element_line(color = "gray33", linetype = 1, size = .2)#,
        # panel.grid.minor.x = element_blank(), 
        # panel.grid.minor.y = element_blank()#,
        # legend.position = "none"
  )

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ # 
# utility functions:
pc <- function(dec) {
  return(round(dec*100, 2))
}
```

---

# Idea

Provide some useful visualizations of common risk information measures as interactive tools (in R and Shiny) for fostering transparent risk communication and teaching purposes.

## Initial Set of Tools

Based on a set of common parameters: population size `N`, prevalence `prev`, sensitivity `sens`, specificity `spec` (or false alarm rate `fa = 1 - spec`).


1. Diagnostic measures based on natural frequencies: 
     a. as 2x2 confusion table (of hits, false alarms, misses, correct rejections)  
     b. as tree with natural frequencies
     
2. Icon array

3. Graphical illustrations of PPV and NPV (given N, prevalence, sensitivity, specificity): 
     a. 2D version: PPV and NPV as function of prevalence (on $x$-axis), with $sens$ and $spec$ as parameters to set.  
     b. 3D version: PPV/NPV as function of $sens$ on $x$-axis, $spec$ on $z$-axis, with $prevalence$ as parameter to set.  

Allow setting and playing with basic parameters or reading in examples from the literature.


## Future Extensions

1. Allow for randomization and sampling

2. Extend to decisions on non-dichotomous distributions (SDT).

3. Extend to the topics of scientific reasoning: _Cell A bias_ and _biased evidence detection_.

---

# Basics 

## Main Parameters

The following variables should be used (and updated) by all parts:

```{r main_parameters, echo = TRUE, eval = TRUE}
## Inputs used in all representations:
# (a) as parameters:
N <- 100 # N in population
prev <- .15 # prevalence in population = p(true positive)
sens <- .85 # sensitivity = p(positive decision | true positive)
spec <- .75 # specificity = p(negative decision | true negative)

# (b) as environment list:
e1 <- list("name" = "Demo",  # name (e.g., HIV, mammography, ...)
           "N" = 100,        # N in population
           "prev" = .15,     # prevalence in population = p(true positive)
           "sens" = .85,     # sensitivity = p(positive decision | true positive)
           "spec" = .75,     # specificity = p(negative decision | true negative)
           "source" = "Source info" # information source (e.g., citation)
           )
```

Alternatively, we could define or read in a set of existing examples:

```{r examples, echo = TRUE, eval = FALSE}
## 1. HIV (a) [Source: Gigerenzer (2013), p. 53]:
name <- "HIVa"
N <- 100000     # women
prev <- 1/10000 # p(HIV)
sens <- 1.0     # p(pos | HIV ) 
spec <- (1 - 5/100000) # 1 - p(fa) = 1 - (pos | no HIV)
source <- "Gigerenzer (2013), p. 53"

## 2. HIV (b) [Source: Gigerenzer (2013), p. 53]:
name <- "HIVb"
N <- 250000     # women
prev <- 1/10000 # p(HIV)
sens <- 1.0     # p(pos | HIV ) 
spec <- (1 - 1/250000) # 1 - p(fa) = 1 - (pos | no HIV)
source <- "Gigerenzer (2013), p. 53"

## 3. Breast cancer  [Source: Gigerenzer (2013), p. 213--221]:
name <- "Breast cancer"
N <- 1000         # women (50yrs, no symptoms, routine mammography screening)
prev <- 10/1000   # p(breast cancer)
sens <- .90       # p(pos | breast cancer ) 
spec <- (1 - .09) # 1 - p(fa) = 1 - (pos | no breast cancer)
source <- "Gigerenzer (2013), p. 213--221"

## 4. Down syndrome  [Source: Gigerenzer (2013), p. 222--226]:
name <- "Down syndrome"
N <- 1000         # women (40yrs, pregnancy)
prev <- 10/1000   # p(Down syndrome)
sens <- .90       # p(pos | Down syndrome ) 
spec <- (1 - .05) # 1 - p(fa) = 1 - (pos | no Down syndrome)
source <- "Gigerenzer (2013), p. 222--226"

## ToDo: Define examples (values and labels) as a data frame (which can be read in or exported).
```

Note that Gigerenzer (2013) provides the _false alarm rate_ (i.e., `1 - sens`), rather than the _sensitivity_ as a basic parameter.

## Data Structures

1. Define some vectors of length `N` based on the (current) main variables:

```{r init_vectors, echo = TRUE, eval = TRUE}
## (0) Get current parameters:
cur.env <- e1

name <- cur.env$name
N <- cur.env$N
prev <- cur.env$prev
sens <- cur.env$sens
spec <- cur.env$spec 
source <- cur.env$source
  
## (1) Determine the truth:
n.true <- round((prev * N), 0)
n.false <- (N - n.true)

truth <- c(rep(TRUE, n.true), rep(FALSE, n.false))
# truth
sum(truth)

## (2) Determine decisions:
n.hi <- round((sens * n.true), 0)  # hits
# n.hi 
n.cr <- round((spec * n.false), 0) # correct rejections
# n.cr
n.mi <- (n.true - n.hi)            # misses
# n.mi
n.fa <- (n.false - n.cr)           # false alarms
# n.fa

decision <- c(rep(TRUE, n.hi), rep(FALSE, n.mi), rep(TRUE, n.fa), rep(FALSE, n.cr))
# decision
# sum(decision)

## (3) Combine truth and decisions:
both.true <-  truth & decision 
both.false <-  !truth & !decision
correct.dec <- both.true | both.false 
length(correct.dec)
sum(correct.dec)/N

## (4) Checks:
N == (n.hi + n.cr + n.mi + n.fa)
length(truth) == length(decision)
sum(truth) == n.hi + n.mi 
sum(decision) == n.hi + n.fa 
sum(both.true) == n.hi 
sum(both.false) == n.cr
sum(correct.dec) == n.hi + n.cr
```

2. Populate the (total) population:

**Goal:** Define a common (global) data structure that is used and updated by all representations.

```{r init_population}
## (1) Initialize the population of individuals:

## (a) Population as a vector of lists (each of which is an individual):
i <- list(tru = NA, # true state of the individual (Boolean) 
          dec = NA, # decision on the individual (Boolean)
          sdt = ""  # category in SDT terms ("hi", "mi", "fa", "cr") 
          )
population <- rep(i, N)

## Populate the population with the values from above:
population$tru <- truth
population$dec <- decision

population$sdt[population$tru & population$dec]   <- "hi"
population$sdt[population$tru & !population$dec]  <- "mi"
population$sdt[!population$tru & population$dec]  <- "fa"
population$sdt[!population$tru & !population$dec] <- "cr"

## (b) Population as a data frame:
population <- data.frame(tru = truth,
                         dec = decision,
                         sdt = NA)
names(population) <- c("truth", "decision", "sdt")

population$sdt[population$tru & population$dec]   <- "hi"
population$sdt[population$tru & !population$dec]  <- "mi"
population$sdt[!population$tru & population$dec]  <- "fa"
population$sdt[!population$tru & !population$dec] <- "cr"

# head(population)
# dim(population)

## (2) Make sdt an ordered factor:
population$sdt <- factor(population$sdt, 
                         levels = c("hi", "mi", "fa", "cr"),
                         # labels = c("hit", "miss", "false alarm", "correct rejection"), # explicit labels
                         labels = c("hi", "mi", "fa", "cr"), # implicit labels
                         ordered = TRUE)
# is.factor(population$sdt)
# levels(population$sdt)

## Checks:
# population$sdt
table(population$sdt)
(sum(population$sdt == "hi") + sum(population$sdt == "cr"))/N # => accuracy
```

Note that defining `population` as a _data frame_ allows immediate access to its variables (columns), cases (rows), and cells. 
Additionally, a rectangular population can be easily displayed, ex- or imported, and sampled. 

# Tools

**Goal:** Provide a range of different perspectives on the same statistical relationships.

## 1. 2x2 confusion table

```{r confusion-table, echo = TRUE, eval = TRUE}
## Compute natural frequencies:

## (1a) Generate from basic parameters (see above):
## 1st row:
# n.true <- round((prev * N), 0)
# n.false <- (N - n.true)
## 2nd row:
# n.hi <- round((sens * n.true), 0)  # hits
# n.cr <- round((spec * n.false), 0) # correct rejections
# n.mi <- (n.true - n.hi)            # misses
# n.fa <- (n.false - n.cr)           # false alarms

## (1b) get from current population:
n.true <- sum(population$tru == TRUE)
n.false <- sum(population$tru == FALSE)

dec.pos <- sum(population$dec == TRUE)
dec.neg <- sum(population$dec == FALSE)

n.hi <- sum(population$sdt == "hi")
n.mi <- sum(population$sdt == "mi")
n.fa <- sum(population$sdt == "fa")
n.cr <- sum(population$sdt == "cr")

# Checks:
N == n.true + n.false
N == dec.pos + dec.neg
n.true  == n.hi + n.mi
n.false == n.fa + n.cr
dec.pos == n.hi + n.fa
dec.neg == n.mi + n.cr

# (2) Define some derived metrics:
acc <- (n.hi + n.cr)/N 
w.sens <- .50
bacc <- (w.sens * n.hi/(n.hi + n.mi)) + ((1 - w.sens) * n.cr/(n.cr + n.fa))
  
ppv <- n.hi / dec.pos # ppv = p(true positive | positive decision)
npv <- n.cr / dec.neg # npv = p(true negative | negative decision)
```

### Showing 2 x 2 confusion table

| Decision: | __true__ | __false__ | Sums: |  
|:----|--:|--:|---:|  
| __positive__ | hi=__`r n.hi`__ | fa=__`r n.fa`__ | `r dec.pos` |  
| __negative__ | mi=__`r n.mi`__ | cr=__`r n.cr`__ | `r dec.neg` |  
| Sums:        | `r n.true`      | `r n.false`     | N = `r N` |  


**Derived metrics:**

- Accuracy: acc = `r pc(acc)`%, bacc = `r pc(bacc)`%,

- Test validities: PPV = `r pc(ppv)`%, NPV = `r pc(npv)`%.


### ToDo

- Make table colored and interactive (using cell values as inputs) 
- Express `prev` as vertical threshold (to be moved from left-to-right to range from 0 to 1)
- Express `sensitivity` as horizontal threshold (...)


## 2. Tree of natural frequencies

1. N = `r N`  
2. n.true = `r n.true` vs. n.false = `r n.false`  
3. (n.hi = `r n.hi` vs. n.mi = `r n.mi`) vs. (n.fa = `r n.fa` vs. n.cr = `r n.cr`)
4. PPV = `r ppv` and NPV = `r npv`

```{r tree-diagram, include = TRUE, fig.width = 10, fig.height = 8}
#install.packages("diagram")
library(diagram)
# demo("flowchart")

## Example 1:
#names <- c("A", "B", "C", "D")
#M <- matrix(nrow = 4, ncol = 4, byrow = TRUE, data = 0)
#plotmat(M, pos = c(1, 2, 1), name = names, lwd = 1, 
#        box.lwd = 2, cex.txt = 0.8, box.size = 0.1, 
#        box.type = "square", box.prop = 0.5)

## Example 2:
#M[2, 1] <- M[3, 1] <- M[4, 2] <- M[4, 3] <- "flow"
#plotmat(M, pos = c(1, 2, 1), curve = 0, name = names, lwd = 1,
#        box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)

## Function to make tree of natural frequencies:
make.nftree <- function(env = cur.env) {
  
  ## (0) Get current parameters:
  name <- cur.env$name
  N <- cur.env$N
  prev <- cur.env$prev
  sens <- cur.env$sens
  spec <- cur.env$spec
  source <- cur.env$source
  
  ## Tree with natural frequencies: 
  names <- c(paste0("N = ", N), # Note: Using global variables (NOT population as argument)
             paste0("true:\n", n.true), 
             paste0("false:\n", n.false), 
             paste0("hits:\n", n.hi), 
             paste0("misses:\n", n.mi),
             paste0("false alarms:\n", n.fa), 
             paste0("correct rejections:\n", n.cr))
  
  M <- matrix(nrow = 7, ncol = 8, byrow = TRUE, data = 0)
  
  M[2, 1] <- "prevalence" # paste0("prevalence = ", as.character(prev)) 
  M[3, 1] <- "(N - true)"
  M[4, 2] <- "sensitivity"
  M[5, 2] <- "(true - hi)"
  M[6, 3] <- "(false - cr)"
  M[7, 3] <- "specificity"
  
  ## plot matrix M:
  pp <- plotmat(M,
                pos = c(1, 2, 4), 
                curve = 0.0,
                name = names,
                box.lwd = 1.5, # radx = 0.1, # rady = 0.05, 
                box.size = .10, 
                box.prop = 0.5,
                box.type = "square", # "circle",
                box.col = "lightyellow", # ... 
                shadow.col = "steelblue4", # "grey25" 
                shadow.size = 0.0, # .005 
                lwd = 1.2,
                cex.txt = .90,
                main = paste0(name, ":\nTree of natural frequencies\n", "(", source, ")")
                )
  
  return(pp)
  
}

nftree <- make.nftree(cur.env)

# typeof(nftree)
# str(nftree)
```


## 3. Icon arrays

Visualize the population, using a common scheme of colors and symbols in all representations.

[To be discussed with Nathaniel.]


## 4. Graphical illustrations of PPV/NPV

### 2D graph

The following graphs illustrate how -- for given test characteristics of `sens` and `spec` -- the metrics `PPV` and `NPV` crucially depend on the _prevalence_ (`prev`) in the population:

```{r ppv-npv-by-prevalence, include = TRUE, fig.width = 10, fig.height = 8}
# sens <- .90 # sensitivity
# spec <- .91 # specificity

# Specify a vector of prevalences:
{
  step.0 <- .10
  prev.0 <- seq(0, 10*step.0, by = step.0)
  step.1 <- .001
  prev.1 <- seq(step.1, 10*step.1, by = step.1)
  step.2 <- .01
  prev.2 <- seq(step.2, 10*step.2, by = step.2)
  step.3 <- .05
  prev.3 <- seq(step.3, 20*step.3, by = step.3)
  step.4 <- .01
  prev.4 <- seq(.90, .90 + 10*step.4, by = step.4)
  step.5 <- .001
  prev.5 <- seq(.990, .990 + 10*step.5, by = step.5)
  
  prev <- sort(unique(c(prev.0, prev.1, prev.2, prev.3, prev.4, prev.5)))
  # prev
  prev.scale <- sort(unique(c(step.0, 5*step.0, step.1, 5*step.1, step.2, 5*step.2, 9*step.0)))
  # prev.scale
}

# (1) Compute PPV and NPV as a function of prev, sens, and spec:
# (A) using Bayes formula:
# as functions:
get.PPV <- function(prev, sens, spec) {
  PPV <- NA # initialize
  num <- (prev * sens)
  den1 <- num
  den2 <- (1 - prev) * (1 - spec)
  PPV <- num / (den1 + den2)
  return(PPV)
}

get.NPV <- function(prev, sens, spec) {
  NPV <- NA # initialize
  num.n <- (1 - prev) * spec
  den1.n <- num.n
  den2.n <- (prev) * (1 - sens)
  NPV <- num.n / (den1.n + den2.n)
  return(NPV)
}

# Use functions to compute current values:
PPV <- get.PPV(prev, sens, spec)
NPV <- get.NPV(prev, sens, spec)

# (B) using natural frequencies:
# (a) choose some population (arbitrary, drops out later)
N <- 10^6 
# (b) 1st row:
n.true <- (prev * N)
n.false <- (N - n.true)
# (c) 2nd row:
n.hi <- (sens * n.true)  # hits
n.mi <- (n.true - n.hi)  # misses
n.cr <- (spec * n.false) # correct rejections
n.fa <- (n.false - n.cr) # false alarms
# (d) derived metrics:
PPV.2 <- n.hi / (n.hi + n.fa) # ppv = p(true positive | positive decision)
NPV.2 <- n.cr / (n.cr + n.mi) # npv = p(true negative | negative decision)

# (C) Verify that both methods yield same results:
all.equal(PPV, PPV.2) 
all.equal(NPV, NPV.2) 

# Plots: 
# plot(PPV)
# m1 <- glm(PPV ~ prev)
# summary(m1)

# (2) Store as data frame:
df <- data.frame(prev, PPV, NPV)
# head(df)

# Reshape into long format:
library("tidyr")
library("dplyr")
df.long <- df %>%
              gather(metric, value, c(PPV, NPV))
# head(df.long)

# factor(df.long$metric)
df.long$metric <- factor(df.long$metric, levels = c("PPV", "NPV")) # ensure factor and level order
# factor(df.long$metric)

sens.spec <- paste0("(sens = ", 100*sens, "%, spec = ", 100*spec, "%)")

# (3) Visualize as a line plot (with ggplot2):
library(ggplot2)
# ggplot(data = df, aes(x = 1:length(prev), y = PPV)) + geom_line()

# A: on linear scale: 
p.lin <- ggplot(data = df.long, aes(x = prev, y = value, group = metric)) +
  geom_line(aes(color = metric), size = 1.2) +
  geom_point(aes(color = metric, shape = metric), size = 2) +
  #geom_smooth(aes(color = metric), size = 1.2) +
  # scale_x_log10(breaks = prev) +
  scale_x_continuous(breaks = seq(0, 1, by = .10)) +
  labs(title = paste0("PPV and NPV by prev ", sens.spec), x = "Prevalence (linear scale)", y = "Probability") +
  scale_color_manual(values = c(col.ppv, col.npv)) +
  # scale_fill_manual(values = c(col.ppv, col.npv), name = "Metric:") +
  my.theme.legend
p.lin

# B: on log scale:
p.log <- ggplot(data = df.long, aes(x = prev, y = value, group = metric)) +
  geom_line(aes(color = metric), size = 1.2) +
  geom_point(aes(color = metric, shape = metric), size = 2) +
  #geom_smooth(aes(color = metric), size = 1.2) +
  # scale_x_log10() +
  scale_x_log10(breaks = prev.scale) +
  # scale_x_continuous(breaks = seq(0, 1, by = .10)) +
  labs(title = paste0("PPV and NPV by prev ", sens.spec), x = "Prevalence (log scale)", y = "Probability") +
  scale_color_manual(values = c(col.ppv, col.npv)) +
  # scale_fill_manual(values = c(col.ppv, col.npv), name = "Metric:") +
  my.theme.legend
# p.log

# C: Combine both plots into one:
library(cowplot)
p.both <- plot_grid(p.lin, p.log,
                    labels = c("A", "B"),
                    nrow = 2, rel_heights = c(.50, .50))
p.both
```

### 3D graph

The following graphs illustrate how -- for a given _prevalence_ (`prev`) in some population -- the metrics `PPV` and `NPV` vary as a function of the _sensitivity_ (`sens`) and the _specificity_ (`spec`) of the test:

```{r ppv-npv-given-prevalence, include = TRUE, fig.width = 10, fig.height = 6}
## Inputs:
prev <- .50 # prevalence (to be set in parameter)
sens.range <- seq(0.0, 1.0, by = .05) # range of sensitivity values 
spec.range <- seq(0.0, 1.0, by = .05) # range of specificity values 

# Compute PPV and NPV for entire matrix of values:
get.matrix <- function(prev, sens, spec, metric) {
  
  # initializing DF (as matrix to store and return results):
  n.rows <- length(sens)
  n.cols <- length(spec)
  matrix <- as.data.frame(matrix(NA, 
                                 nrow = n.rows, 
                                 ncol = n.cols)) 
  names(matrix) <- sens 
  
  # loop through all rows and columns of pc.matrix: 
  for (row in 1:n.rows) {
    for (col in 1:n.cols) {
      
      # Compute the needed model DV for the current cell value:
      cell.val <- NA 
      
      if (metric == "PPV") {cell.val <- get.PPV(prev, sens[row], spec[col])} # compute PPV
      if (metric == "NPV") {cell.val <- get.NPV(prev, sens[row], spec[col])} # compute NPV
      
      # Store results:
      matrix[row, col] <- cell.val 
      
    }
  }
  
  return(matrix)
  
}

# Use function to compute current values:
PPV.mat <- get.matrix(prev, sens.range, spec.range, metric = "PPV")
NPV.mat <- get.matrix(prev, sens.range, spec.range, metric = "NPV")

### Visualize as 3D graph (via persp): 
x <- sens.range
y <- spec.range
z.ppv <- as.matrix(PPV.mat)
z.npv <- as.matrix(NPV.mat)

z.lim <- c(0, 1) # range of z-axis

# persp(x, y, z.ppv)
# persp(x, y, z.ppv)

# Set basic graph parameters:
my.theta <- 40 # horizontal viewing angle (higher values: more rotation)
my.phi <- 10   # vertical viewing angle (higher values: higher viewpoint)
my.expand <- .9 # values < 1 shrink expansion in z-direction
my.d <- 1.2 # values > 1 lessen perspective effect 

# Plot both plots (next to each other):
{
par(mfrow=c(1,2)) # Combine 2 plots in 1 row x 2 columns.
par(bg = "white")

p.ppv <- persp(x, y, z.ppv, 
               theta = my.theta, phi = my.phi,  d = my.d, expand = my.expand, col = col.ppv, ltheta = 200, shade = 0.10, 
               ticktype = "detailed", xlab = "sens", ylab = "spec", zlab = "PPV", zlim = z.lim, 
               main = paste0("PPV (prevalence = ", 100 * prev, "%)"))

p.npv <- persp(x, y, z.npv, 
               theta = my.theta, phi = my.phi,  d = my.d, expand = my.expand, col = col.npv, ltheta = 200, shade = 0.10, 
               ticktype = "detailed", xlab = "sens", ylab = "spec", zlab = "NPV", zlim = z.lim, 
               main = paste0("NPV (prevalence = ", 100 * prev, "%)"))

par(mfrow=c(1,1)) # Remove special settings.
}
```


# Related Topics

Some topics for future extensions:

## Cell A bias and biased sampling of evidence

### Contingency table

Assume the following:

- 2 independent probabilities for features $A$ and $B$ that suggest a causal link (e.g., 
  [$A$ = nuclear power station and $B$ = cancer] or 
  [$A$ = aluminum deodorant and $B$ = cancer] etc.

- $A$ and $B$ are statistically independent (not causally related)

```{r causal_2x2, echo = TRUE, eval = TRUE}
A.cat <- c("alu", "no_alu") # A is binary
B.cat <- c("cancer", "healthy") # B is binary
pA <- .66
pB <- .10

# Simulation:
N <- 1000
A <- sample(x = A.cat, size = N, replace = TRUE, prob = c(pA, 1 - pA))
B <- sample(x = B.cat, size = N, replace = TRUE, prob = c(pB, 1 - pB))

# Contingency table:
BxA <- table(B, A)
BxA

# sums and percentages:
colSums(BxA)
rowSums(BxA)
pc(BxA/sum(BxA)) 

# 1st row percentages per column (pB):
pc(BxA[1,1]/sum(BxA[ , 1]))
pc(BxA[1,2]/sum(BxA[ , 2]))

# 1st col percentages per row (pA)
pc(BxA[1,1]/sum(BxA[1, ]))
pc(BxA[2,1]/sum(BxA[2, ]))

# Focus on row 1 suggests causal relationship A->B (which is simply due to pA). 


# # Classify cases:
# X <- rep(NA, N)
# for (i in 1:length(A.cat)) {
#  for (j in 1:length(B.cat)) {
#    
#    X[A == A.cat[i] & B == B.cat[j]] <- (i-1)*length(B.cat) + j
# } 
# }
# table(X)
```

**ToDo:** Add measures of relatedness and causality (e.g., odds ratios, Matthew's correlation, etc.)

### Biased sampling

Now assume that we do not have access to the overall contingency table, but only instances of individual cases in memory, from which we _sample_ (or to which we _selectively attend_).

- Unbiased sampling would approximate the true contingency table.

- However, focusing on particular dimensions (of $A$ or $B$) would yield biased tables and estimates. 


# Conclusion

## Important Insights

Materials for conclusions and possible quizzes:

- The sensitivity and specificity of a diagnostic test are insufficient to determine its PPV and NPV.

- Both PPV and NPV crucially depend on the specific _prevalence_ of a particular person. 
Thus, always use an appropriate reference group to specify the risk of a particular person as accurately as possible.

- PPV and NPV show _opposite_ trends (as a function of prevalence): When one is high, the other one tends to be low (and vice versa).

     - with increasing prevalence: PPV increases.
     
     - with increasing prevalence: NPV decreases.


## Recommendations for the General Population

- A risk-literate person should be aware that any diagnostic test (or clinical judgment) has 4 possible outcomes: 2 possible correct and 2 possible incorrect judgments.

- Whenever receiving a (positive or negative) test result, a good question is: 
What is the _probability_ that this result is _wrong_?

    - Whenver receiving a positive (undesirable) test result: What is the PPV vs. p(false alarm) for me (or a random person from my reference group)?

    - Whenever receiving a negative (desirable) test result: What is the NPV vs. p(miss) for me (or a random person from my reference group)?


## Recommendations for Risk Professionals (e.g., Medical Health Experts)

More challenging conclusions and demands:

- All risk professionals should know how to determine and evaluate evidence for and against treatments (as a function of all 4 cells, rather than committing cell A bias). 

- All risk professionals should be able to define and distinguish between _sensitivity_ and _PPV_ (as opposite conditional probabilities), as well as _specificity_ and _NPV_ (sic).

- All risk professionals should be able to use _contingency tables_, _natural frequency trees_, and _icon arrays_ to solve Bayesian problems (of conditional probability).

--- 

[``r fileName`` updated on `r Sys.time()` by [hn](http://neth.de/).]

<!-- +++ finis eof -->